---
title: "Denominators"
subtitle: Working with OFM's SADE Products
author: "Daniel Casey"
format:
  revealjs:
    theme: solarized
editor: visual
---

## Small Area, Big Heart

Small Area Demographic Estimates (SADE) are a set of modeled population estimates produced by WA OFM

- Census Block

- 62 Race/Ethnicity Combinations

- Sex

- 2000 - 2025

They are the main source of denominators for WA.

## You've Got Options

| [WASP]{style="font-size: 1.25em;"} | [CHAT]{style="font-size: 1.25em;"} |
|------------------------------------|------------------------------------|
| Locally hosted (requires downloading some files) | A familiar friend |
| **AOIC Estimates** | Pointy-clicky interface |
| **R interface and machine readable outputs** | Remotely hosted and maintained |
| No ZIP code data (yet) |  |
| Pointy-clicky interface available |  |
| Customizable\* |  |

## Ask Nicely

Email [ofmforecasting\@ofm.wa.gov](mailto:ofmforecasting@ofm.wa.gov) to request access/sign the EUA.

- [2000 - 2009](https://data.wa.gov/en/demographics/Small-Area-Demographic-Estimates-2000-2009/g3gh-r5g7/about_data), [2010 - 2019](https://data.wa.gov/en/demographics/Small-Area-Demographic-Estimates-2010-2019/ivkq-ti7d/about_data), [2020 - Present](https://data.wa.gov/en/demographics/Small-Area-Demographic-Estimates-2010-2019/ivkq-ti7d/about_data), [Geographic Crosswalk](https://data.wa.gov/demographics/OFM-Geographic-Crosswalk/pvty-6zcu/about_data)

- Downloads can take an hour per dataset, but you can pre-filter if you want

Review the [Onboarding Guide](https://github.com/OFMPop/ofm_opendata_portal_info/blob/main/guides/partner_onboarding.md) for more information on accessing and downloading the data and the [Data Interaction Guide](https://github.com/OFMPop/ofm_opendata_portal_info/blob/main/guides/data_interaction.md) for accessing subsets.

## Processing ... please quack

With the data downloaded, and some file paths set in the script or in .Renviron, the [process_data.R](../Scripts/process_data.R) script will load, reformat, and save the input CSVs in to a DuckDB database.

<br>

DuckDB is a great analytical tool:

- Local in-memory/on-disk database

- Very fast and good at analytical workflows

- Good flavor of SQL and interoperability with R (see also duckplyr or dbplyr)

## Fetch results

The [fetch_pop](../Scripts/Custom_Functions/fetch_data_localdb.R) function extracts processed population data from the DB.

```{r, include=TRUE, echo=TRUE}

# Some code to set the environment
library('data.table')
library('DBI')
library('duckdb')
library('glue')
library('gt')
source('../Scripts/Custom_Functions/fetch_data_localdb.R')
dbpath = file.path(Sys.getenv("OUTPUT_FILEPATH"), 'popdb.duckdb')

# connect to db
db_ro = dbConnect(duckdb::duckdb(), dbpath, read_only = T)

totab = function(x) gt(x) |> tab_options(table.width = pct(100))

```

```{r}
# From: https://gt.albert-rapp.de/quarto_gt
#library(knitr)
#knit_print.gt <- function(x, ...) {
#  stringr::str_c(
#    "<div style='all:initial';>\n", 
#    gt::as_raw_html(x), 
#    "\n</div>"
#  ) |> 
#    knitr::asis_output()
  
#}
#registerS3method(
#  "knit_print", 'gt_tbl', knit_print.gt, 
#  envir = asNamespace("gt") 
#  # important to overwrite {gt}s knit_print
#)
# 
```

## Age Options

```{r age-opts, include=TRUE, echo=TRUE}
age_opts = dbGetQuery(db_ro, 'select * from age_tab')
head(age_opts) |> totab()
```

## Race/Ethnicity Options

```{r race-opts, include=TRUE, echo=TRUE}
race_opts = dbGetQuery(db_ro, 'select * from re_grid')
head(race_opts) |> totab()
```

## Geographic Options

```{r geog-opts, include=TRUE, echo=TRUE}
geog_opts = dbGetQuery(db_ro, 'show tables') |>
  subset(! name %in% c('age_tab', 're_grid', 'geog_xw'))

geog_opts$name

```

You can refer to OFM's geography crosswalk to go from FIPS code to something more human readable.

## Example #1

::: {style="font-size: 0.7em;"}
Population of Kitsap County: Males 25-34, 2021 and 2025, White alone (incl. Hispanic)
:::

```{r examp1, include=TRUE, echo=TRUE}
fetch_pop(
  dbpath = dbpath, geog_level = 'county', geog = 53035, year = c(2021, 2025),
  age_col = 'age_11g', age = '25-34', gender = 'Male', raceeth_col = 'race_code', 
  raceeth = c(10000), groups = c('Year', "Race/Eth") ) |> totab()

```

```{r, include = TRUE, echo=TRUE}

fetch_pop(
  dbpath = dbpath, geog_level = 'county', geog = 53035, year = c(2021, 2025),
  age_col = 'age_11g', age = '25-34', gender = 'Male', raceeth_col = 'race6', 
  raceeth = 'White Only', groups = c('Year', "Race/Eth")) |> totab()
```

## A brief aside on AOIC

Just because AOIC estimates exist and are available does not mean they mean what you think they mean (or what you want them to mean).

- Mis-classification

  - Joshi et al (2018): Vital statistics data alone underestimated overdose mortality among AIAN people by \~40%

- Difference in ascertainment method between data versions and data systems (e.g. Census 2010 -\> Census 2020, SADE vs. Vital Registration)

- Data processing protocols (e.g. imputation from ethnicity)

## Example #2

::: {style="font-size: 0.7em;"}
Population of Seattle School District: Asian (alone or in-combination) by 5 year age groups under 18
:::

```{r examp2, include = TRUE, echo=TRUE}
fetch_pop(dbpath, geog_level = 'schooldist', geog = '5307710', year = 2025,
          age_col = 'age_5yr', age = c('5-9','10-14', '15-17', '18-19'),
          gender = 'All', raceeth_col = 'race_as', raceeth = 1,
          groups = c('Year', 'Age')) |> totab()

fetch_pop(dbpath, geog_level = 'schooldist', geog = '5307710', year = 2025,
          age_col = 'age_5yr', age = c('5-9','10-14', '15-17', '18-19'),
          gender = 'All', raceeth_col = 'AIC', raceeth = 'as',
          groups = c('Year', 'Age')) |> totab()

```

## Example #3

::: {style="font-size: 0.7em;"}
Population for King County, White alone or in-combination, and AIAN alone or in-combination. Both non-Hispanic.
:::

```{r examp3, include = TRUE, echo=TRUE}
fetch_pop(dbpath, geog_level = 'county', geog = '53033', year = 2025,
          raceeth_col = 'AIC-NH', raceeth = c('aian', 'wht'),
          groups = c('Year')) |> totab()
```

## Example #4

::: {style="font-size: 0.7em;"}
Return the SQL code of a query that does not use AIC/AIC-NH for race (that will instead return multiple queries)
:::

```{r examp4, echo = TRUE, include = TRUE}
fetch_pop(
  dbpath = dbpath, geog_level = 'county', geog = 53035, year = c(2021, 2025),
  age_col = 'age_11g', age = '25-34', gender = 'Male', raceeth_col = 'race6', 
  raceeth = 'White Only', groups = c('Year', "Race/Eth"), return_query = T)[[1]]
```

## App App Baby

The old [PopPIE R-Shiny app](../Scripts/app.R) has been retooled to use the DuckDB created via process_data.R script. Note: you may need to install some extra packages. [See the Github repo for more](github.com/WA-LHJ-Analytic-Resources/WA-OFM-SADE-Population-Estimates).

## Git is your friend and you must use it

Getting this code created and working was collaborative effort. We used git and GitHub to manage the code release.

<br>

If you write code, use git. It eliminates the need to make `analysis_final_final_final`. Instead you just make a branch! You can also rollback to older versions of the code if you need to. Its great. use it. Use it. USE IT.

<br>

[A good example of how to use GitHub for collaboration.](https://github.com/WA-LHJ-Analytic-Resources/WA-OFM-SADE-Population-Estimates/pull/8)
